---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# **seine**: Semiparametric Ecological Inference <a href="https://corymccartan.com/seine/"><img src="man/figures/logo.svg" align="right" height="144" /></a>

<!-- badges: start -->
[![R-CMD-check](https://github.com/CoryMcCartan/seine/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CoryMcCartan/seine/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Ecological inference (EI) is the statistical problem of learning individual-level
associations from aggregate-level data.
Without certain *identifying assumptions* and proper *estimation methods*,
researchers can easily draw incorrect conclusions from aggregate data, finding
patterns where none exist, missing important individual-level patterns, or
even concluding an effect runs in the opposite direction than it actually does.
This is known as the *ecological fallacy*.

The **seine** package allows researchers to perform modern ecological inference
quickly, accurately, and transparently.

- *Double/debiased machine learning* allows for controlling for confounding 
  covariates, which increases the plausibility of identifying assumptions.
- *Sensitivity analysis* and *benchmarking* let researchers understand how
  violations of their assumptions will affect results.
- A *tidy interface* makes the package modular and easy to use, and works well
  with pipe-based workflows.
- *Minimal dependencies* and efficient estimation routines keep everything 
  *fast* and lightweight.

<h2><span style="text-color: red">This software is experimental</span></h2>

**seine** and the underlying statistical methods are under active development.
The package is not currently recommended for non-expert use.

## Installation

You can install the development version of **seine** with

``` r
remotes::install_github("CoryMcCartan/seine")
```

## Example

The package contains county-level data from the 1968 U.S. presidential election
for southern states, where Hubert Humphrey (Democrat) ran against Richard Nixon
(Republican) and George Wallace (independent; segregationist).

In this data, one might be interested in estimating vote choice by race.
To do so, **seine** provides both a formula interface and a tidy interface
through a new `ei_spec()` object. Here, we'll demonstrate the latter.

```{r}
library(seine)
data(elec_1968)

spec = ei_spec(
    elec_1968, 
    predictors = vap_white:vap_other,
    outcome = pres_dem_hum:pres_ind_wal, 
    total = pres_total,
    covariates = c(farm:educ_coll, inc_00_03k:inc_25_99k)
)

m = ei_ridge(spec)
rr = ei_riesz(spec, penalty = m$penalty)

ei_est(regr = m, riesz = rr, data = spec, conf_level = 0.95)
```

This workflow is explained in more detail in the [package vignette](./articles/seine.html), along
with demonstrations of data preprocessing and sensitivity analysis.

