---
title: "Local estimates for ecological inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Local estimates for ecological inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Global ecological inference estimates describe average behavior across all units
in the data.
Sometimes, however, estimates for each individual unit (precinct, county, or
other geography) are of interest: to map spatial variation, to examine outliers,
or to inform decisions at the local level.
**seine** provides two approaches for this.
`ei_bounds()` computes guaranteed-valid partial identification bounds for each
unit, relying only on the accounting identity and known limits on the outcome.
`ei_est_local()` produces point estimates and confidence intervals under the
Conditional Average Representativeness (CAR) assumption and a model for
within-unit variation, using a fitted regression model.
The former bounds are also sometimes called the Duncan-Davis bounds.

# Setting up the analysis

We will use the `elec_1968` data included in the package, which contain
county-level election returns from Southern states in the 1968 U.S.
presidential election.
We want to estimate the individual-level association between race and
presidential vote choice for each county.

```{r setup}
library(seine)
data(elec_1968)
```

For the bounds, only the data are needed.
For `ei_est_local()`, we also need a fitted regression model, so we set up an
`ei_spec` object and call `ei_ridge()`.

```{r}
spec = ei_spec(
    elec_1968,
    predictors = vap_white:vap_other,
    outcome = pres_dem_hum:pres_abs,
    total = pres_total,
    covariates = c(state, pop_city:pop_rural, farm:educ_coll, inc_00_03k:inc_25_99k),
    preproc = function(x) {
        x = model.matrix(~ 0 + ., x) # convert factors to dummies
        bases::b_bart(x, trees = 200)
    }
)

m = ei_ridge(spec)
```

See the main vignette (`vignette("seine")`) for a full walkthrough of the
estimation workflow.

# Partial identification bounds

The simplest local quantities to compute are partial identification bounds:
for each county and each predictor-outcome combination, the range of values
for the local estimand $\beta_{gj} = \Pr(Y = j \mid X = x_g, G = g)$ that is
consistent with the observed aggregate data.
These bounds require no modeling assumptions beyond the accounting identity and
the specified bounds on the outcome.

`ei_bounds()` takes an `ei_spec` object (or a formula) and returns a data frame
with one row per unit per predictor-outcome combination.

```{r}
bnds = ei_bounds(spec, bounds = c(0, 1))
head(bnds)
```

The `min` and `max` columns give the sharp bounds for each unit.
The `weight` column is the product of the predictor proportion and the total,
which is useful for aggregating.
When the bounds are wide for many units, this indicates a fundamentally
difficult identification problem, regardless of modeling choices.

To aggregate the bounds across all units to produce bounds on the global
estimands, pass `global = TRUE`.

```{r}
ei_bounds(spec, bounds = c(0, 1), global = TRUE)
```

The `as.array()` method reformats the output as a three-dimensional array with
dimensions units by predictors by outcomes, which can be convenient for further
analysis.

```{r}
bnds_arr = as.array(bnds)
dim(bnds_arr) # rows x predictors x outcomes
```

Bounds on contrasts, such as the difference in vote share between White and
Black voters, can be computed directly by passing a `contrast` argument.

```{r}
ei_bounds(spec, bounds = c(0, 1), contrast = list(predictor = c(1, -1, 0)))
```

# Estimating the local covariance

Point estimates and confidence intervals from `ei_est_local()` require a model
for how the local estimands vary around their conditional mean within each unit.
This variation is captured by the `b_cov` argument, a covariance matrix for
the local estimands $\beta_{gj}$ around the regression prediction.

`ei_local_cov()` estimates this covariance from the data, under the CAR
assumption and an additional homoskedasticity condition.
Concretely, it fits a ridge regression of the empirical second moments of the
regression residuals on the cross-products of the predictor variables, and uses
the polarization identity to recover the covariance for each predictor-outcome
pair.
A small amount of shrinkage towards the global residual covariance is applied,
controlled by `prior_obs`.

```{r}
b_cov = ei_local_cov(m, spec)
round(b_cov, 3)
```

The rows and columns are ordered by predictor within outcome, i.e.
(Y1|X1, Y1|X2, ..., Y2|X1, Y2|X2, ...).
Large diagonal entries indicate high within-unit variation in the local
estimands, relative to the regression predictions.
Off-diagonal entries reflect the extent to which, within a county, vote choices
across racial groups tend to move together.

# Local point estimates

With the regression model and a covariance structure in hand, `ei_est_local()`
projects the regression predictions onto the accounting constraint to produce
local estimates that exactly satisfy the ecological identity within each county.
The function also computes valid confidence intervals using Chebyshev's
inequality or, under a unimodality assumption, the tighter unimodal bound.

The `b_cov` argument controls the assumed covariance structure.
There are three common choices:

- `b_cov = ei_local_cov(m, spec)`: the **data-estimated covariance**, which
  uses the structure estimated above. *This is the recommended choice*.
- `b_cov = 0`: the **orthogonal model**, which assumes no correlation in the
  local estimands across predictors within a unit. When the estimated `b_cov`
  is implausible, this can be an acceptable choice.
- `b_cov = 0.95` (or another value near 1): the **neighborhood model**, which
  assumes nearly perfect positive correlation across predictors. This is
  appropriate when counties are relatively homogeneous and the local estimands
  are expected to vary together. However, it leads to quite narrow confidence
  intervals, which may not cover the truth in practice.

We fit all three and compare the average width of the resulting confidence
intervals.

```{r}
e_rcov = ei_est_local(m, spec, b_cov = b_cov, bounds = c(0, 1), sum_one = TRUE)
e_orth = ei_est_local(m, spec, b_cov = 0,    bounds = c(0, 1), sum_one = TRUE)
e_nbhd = ei_est_local(m, spec, b_cov = 0.95, bounds = c(0, 1), sum_one = TRUE)

c(
    estimated = mean(e_rcov$conf.high - e_rcov$conf.low),
    orthogonal = mean(e_orth$conf.high - e_orth$conf.low),
    neighborhood = mean(e_nbhd$conf.high - e_nbhd$conf.low)
)
```

The orthogonal model tends to produce the widest intervals.
The neighborhood model can be narrower when the predictor proportions within a
county are nearly uniform, because the accounting constraint then binds tightly.
The estimated covariance typically lies between the two extremes, and as
mentioned above is the recommended default.

Setting `sum_one = TRUE` enforces the constraint that the local vote shares
across candidates sum to one within each racial group, which is the correct
restriction for these data.
The `bounds = c(0, 1)` argument truncates estimates to the unit interval and
caps the standard errors implied by the width of the bounds.

To examine results for a specific county, we can filter the output.

```{r}
subset(e_rcov, .row == 1)
```

The `as.array()` method provides a convenient view of the point estimates as a
three-dimensional array.

```{r}
head(as.array(e_rcov)[, , "pres_dem_hum"])
```

Finally, taking a weighted average of the local estimates against the `weight`
column reproduces the global estimate from `ei_est()`.

```{r}
grps = split(e_rcov, ~ predictor + outcome)
sapply(grps, function(g) weighted.mean(g$estimate, g$weight))
```

# References

McCartan, C., & Kuriwaki, S. (2025+). Identification and semiparametric
estimation of conditional means from aggregate data.
Working paper [arXiv:2509.20194](https://arxiv.org/abs/2509.20194).

Chernozhukov, V., Cinelli, C., Newey, W., Sharma, A., & Syrgkanis, V. (2024).
Long story short: Omitted variable bias in causal machine learning
(No. w30302). *National Bureau of Economic Research.*


<small>
This vignette was originally produced by a large language model, and then reviewed and edited by the package authors.
</small>