---
title: "Analyzing aggregate data with **seine**"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing aggregate data with 'seine'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Ecological inference (EI) is the statistical problem of learning individual-level
associations from aggregate-level data.
EI commonly arises when two datasets are joined using a shared geographic
identifier, and when individual data are not released for privacy reasons.
To take some recent examples from the *New York Times*:

- [Estimating COIVD vaccine uptake by political beliefs](https://www.nytimes.com/interactive/2021/04/17/us/vaccine-hesitancy-politics.html)
- [Understanding which demographics supported a progressive mayoral candidate](https://www.nytimes.com/interactive/2025/06/25/nyregion/nyc-mayor-election-results-map-mamdani-cuomo.html)
- [Evaluating the differential impact of tariffs on political partisans](https://www.nytimes.com/interactive/2025/03/15/business/economy/tariffs-trump-maps-voters.html)

EI is also used in public health and epidemiology, and is widely applied in
litigation under the federal Voting Rights Act of 1965 (VRA) to establish the
presence of racially polarized voting.

# Avoiding the ecological fallacy

# Preparing data

```{r setup}
library(seine)
```

# Estimation

## The big picture

## Fitting the regression and Riesz representer

## DML for ecological estimates

# Sensitivity Analysis
