---
title: "Sensitivity analysis for ecological inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis for ecological inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the sensitivity analysis tools in **seine**,
using the `elec_1968` data on county-level voting in Southern states in the
1968 U.S. presidential election.
Sensitivity analysis is essential for ecological inference (EI) because all EI
methods rely on an untestable identifying assumption---here, Conditional
Average Representativeness, or CAR---that is unlikely to hold exactly in practice.
The tools in **seine** are based on a nonparametric sensitivity framework
developed by Chernozhukov et al. (2024).

# Setting up the analysis

We begin by loading the 1968 election data, and defining an `ei_spec` object
that records the outcome, predictor, covariate, and total-count columns,
following the setup from `vignette("seine")`.
We use a BART basis expansion for nonparametric covariate adjustment, which
we strongly recommend to avoid dependence on linearity assumptions.

```{r setup}
library(seine)
data(elec_1968)

spec = ei_spec(
    elec_1968,
    predictors = vap_white:vap_other,
    outcome = pres_dem_hum:pres_abs,
    total = pres_total,
    covariates = c(state, pop_city:pop_rural, farm:educ_coll, inc_00_03k:inc_25_99k),
    preproc = function(x) {
        x = model.matrix(~ 0 + ., x) # convert factors to dummies
        bases::b_bart(x, trees = 200)
    }
)
```

We fit the regression model with `ei_ridge()` and the Riesz representer with
`ei_riesz()`, then combine them with `ei_est()` to estimate vote choice by race
using double machine learning (DML).
We focus on the *contrast* between White and Black voters, which is a direct
measure of racially polarized voting.
See the main vignette (`vignette("seine")`) for a full walkthrough of this
estimation workflow.

```{r}
m = ei_ridge(spec)
rr = ei_riesz(spec, penalty = m$penalty)

est = ei_est(m, rr, spec, contrast = list(predictor = c(1, -1, 0)), conf_level = FALSE)
print(est)
```

While we normally do not recommend setting `conf_level = FALSE` to suppress confidence intervals, here we do, so that the output can more easily fit on the screen.
If confidence intervals are present in `est`, they will be adjusted by the senstivity analysis below.

# Sensitivity analysis

The estimates above rest on the CAR assumption: that, conditional on the
observed covariates, individual vote choice is independent of the individual's
race.
In practice, this assumption is unlikely to hold exactly, as there may be
unobserved confounders.
**seine** provides a number of tools to evaluate how sensitive the results are
to violations of that assumption.

The sensitivity framework considers the relationship between an unobserved
confounding variable and (i) the outcome and (ii) the Riesz representer,
measured in terms of partial $R^2$ values (`c_outcome` and `c_predictor`,
respectively).
Stronger relationships indicate more confounding and therefore more potential
bias in the original estimates.

The `ei_sens()` function provides a simple interface to this framework.
Users provide values for the sensitivity parameters, and a bound on the absolute
bias is returned.
In the following example, we investigate the effect of an omitted confounder
that explains 50% of the residual variation in the outcome and 20% of the
variation in the Riesz representer.

```{r}
ei_sens(est, c_outcome = 0.5, c_predictor = 0.2)
```

We can also work backwards and ask what one of the sensitivity parameters would
have to be in order to produce a certain amount of bias.
For example, if we assumed a worst-case scenario where the confounder explains
the entire outcome (`c_outcome = 1`), we can ask how strongly that confounder
would need to be related to the Riesz representer to produce a bias of up to
5pp.

```{r}
ei_sens(est, c_outcome = 1, bias_bound = 0.05)
```

For most predictors and outcomes, the answer is not very much!

## Benchmarking

The `c_outcome` parameter is relatively easy to understand, but `c_predictor`
is more difficult to interpret (though see the methodology paper for more
discussion).
To help understand plausible values of these parameters, we can conduct a
**benchmarking analysis** that treats each of our *observed* covariates in turn
as a hypothetical *unobserved* confounder, and calculates the implied values of
the sensitivity parameters.

```{r}
bench = ei_bench(spec, contrast = list(predictor = c(1, -1, 0)))

subset(bench, outcome == "pres_rep_nix")
```

```{r include=FALSE}
fmt_pp = \(x) paste0(format(100*x, digits=2), "pp")
est_pt = subset(est, outcome=="pres_rep_nix")$estimate 
bb = subset(bench, covariate == "state" & outcome == "pres_rep_nix")
est_bias = subset(ei_sens(est, bb$c_outcome, bb$c_predictor), outcome == "pres_rep_nix")$bias_bound
est_max = max(abs(bench$est_chg))
```

The table above shows the benchmark values for each covariate for the racially polarized Nixon vote estimand.
The `confounding` column is an additional component of the sensitivity analysis
that is discussed in the paper; the default value is 1, which is a conservative
worst-case bound.
The benchmark values here show that `state` is far and away the strongest
observed confounder, whose inclusion changes the estimate by `r fmt_pp(est_max)`.
If the unobserved confounders were as strong as `state`, we might expect a
significant amount of bias, as we will see next.

## Bias contour plot

Rather than perform this sensitivity analysis on a single set of sensitivity
parameters, we can run it across all combinations of parameter values, and
visualize the results on a **bias contour plot.**
We can further overlay the benchmarking values to help interpret the results.

```{r}
#| fig.alt: >
#|   Bias contour plot for the racially polarized Nixon vote
sens = ei_sens(est) # the default evaluates on a grid of parameters
plot(sens, "pres_rep_nix", bench = bench, bounds = c(-1, 1))
```

The contour lines indicate how much bias could result from an unobserved
confounder with the specified sensitivity parameters.
The blue dashed contours correspond to bias of 1, 2, and 3 standard errors.
This is a helpful value to compare against, because bias of that size
corresponds to a predictable drop in coverage rates of confidence intervals.
For example, bias of 1 standard error means that a confidence interval with
95% nominal coverage will actually have coverage of only around 80%.

The red asterisks indicate the benchmark values for each covariate.
Most are clustered in the lower-left corner and can't be distinguished. In
contrast, the benchmark for `state` shows that an unobserved confounder of that
strength could lead to bias of around `r fmt_pp(est_bias)`, which is substantial
compared to the estimate itself, which is `r fmt_pp(est_pt)`.

## Robustness value

Finally, it can be helpful to summarize the sensitivity analysis by a single
number.
The `ei_sens_rv()` function calculates the **robustness value**, which measures
the minimum strength of an unobserved confounder that would lead to a bias of
a given amount.
Here, we consider bias sufficient to eliminate any evidence of racially
polarized voting, i.e., bias equal to the estimated difference between White
and Black voters.

```{r}
ei_sens_rv(est, bias_bound = estimate)
```

The robustness value (one for each predictor/outcome combination) is
relatively small for Wallace's vote share, indicating low robustness (high
sensitivity).
In particular, it is far smaller than the amount of confounding benchmarked by
the observed `state` variable.
For Humphrey and Nixon's vote shares, however, the robustness values are larger,
indicating more confidence in the finding of racially polarized voting for those
candidates.

As with any single-number summary, it is important to consider sensitivity
beyond the single value, by using the contour plot and the benchmarking
analysis.

# References

McCartan, C., & Kuriwaki, S. (2025+). Identification and semiparametric
estimation of conditional means from aggregate data.
Working paper [arXiv:2509.20194](https://arxiv.org/abs/2509.20194).

Chernozhukov, V., Cinelli, C., Newey, W., Sharma, A., & Syrgkanis, V. (2024).
Long story short: Omitted variable bias in causal machine learning
(No. w30302). *National Bureau of Economic Research.*
